### 2014-11-30 paper.tex
Written introduction section.

\documentclass{report}

\usepackage[style=authoryear,backend=biber]{biblatex}

\title{The Folly of Passwords}
\author{Ross Fenning}

\bibliography{references}

\begin{document}

\maketitle
\tableofcontents

\chapter{Introduction}

The use of passwords for authentication is certainly commonplace.
% TODO: How many passwords?
It is standard practice within small and large enterprises alike
to require employees to authenticate before using desktop
computers, mobile devices or centralised services (e.g.
as company email, corporate intranet or web-based self-service
systems for tasks such as booking annual leave).

These authentication requirements are largely driven by
several desires:

\begin{itemize}
  \item to prevent use of systems by non-employees;
  \item to maintain an audit trail of employees' usage;
  \item to protect company secrets and intellectual property;
  \item to safeguard the personal information of customers and employees; and
  \item to prevent economic and reputational damage from would-be attackers.
\end{itemize}

Additional security in any system normally comes with additional costs
in terms of usability for users and in financial terms for the organisation
that must implement any security processes and policies. However,
whilst many of the benefits and requirements listed above should justify
these costs, they are notably gains (or prevention of losses) for
the organisation itself more than they are gains for the individual
employees.

In chapter~\ref{chapter:costs}, we will explore the
cost-benefit trade-offs more closely and demonstrate that the status
quo unreaslistically transfers costs to employees to support gains
for the organisations.

Passwords are not the only approach to verifying a user's identity.
The methods by which a user can be authenticated are known to fall
into three categories -- or \emph{factors} -- of authentication
\parencite{council2005authentication}:

\begin{itemize}
  \item something the user \emph{knows}, such as a password, a PIN or
    personal information (e.g. mother's maiden name);
  \item something the user \emph{has}, such as a passport, a mobile device
    a paper ticket; and
  \item something the user \emph{is}, such as their DNA sequence,
    their retinal pattern, their fingerprint.
\end{itemize}

It should be noted that some methods can be argued as spanning
over two factors. For example, a credit card is something
a bank customer possesses, but the card number is something
an attacker could learn without the owner losing ownership
of the card. In the context of many online purchase, credit
card \emph{details} arguably become something a user \emph{knows}.

User authentication that requires two differing factors
of authentication (known as \emph{two-factor authentication}. There
has been a rise in two-factor authenication with companies
such as Google, Amazon, Microsoft, Paypal and Twitter having enabled
two-factor authentication for user accounts.

The mainstream adoption of authentication methods other than
passwords suggests increased ease in enterprises adoption means
of verifying users either instead of or in addition to
existing password setups.

In chapter~\ref{chapter:alternatives}, these alternatives will be
compared and assessed for their ease of adoption within a business.

\chapter{Problems with Passwords}
\label{chapter:costs}

\section{The Costs of Passwords}
\section{Password Strength}
\section{Expiry}

\chapter{Comparative Analysis of Password Alternatives}
\label{chapter:alternatives}

\chapter{Conclusions and Recommendations}

\chapter{Further Work}

\printbibliography

\end{document}

### 2014-11-30 wordcount.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+perl texcount.pl "$@" | grep 'Words in text' | tail -n 1 | sed 's/.*: //g'

### 2014-11-30 paper.tex
Written a lot of chapter 2.

@@
 
-\chapter{Problems with Passwords}
+\chapter{The Cost of Passwords}
 \label{chapter:costs}
 
-\section{The Costs of Passwords}
+The username-password pair is arguably a \emph{de facto} standard
+authentication mechanism employed by IT and security departments
+within many businesses. The default install of any operating
+system (Linux, Mac OS X, Windows, etc.) treats all user accounts
+as consisting minimally of these pair of credentials and a
+username-password login prompt is frequently the first
+screen to greet users when a newly-installed OS is booted
+for the first time.
+
+In this chapter, we will explore the burden passwords place
+on users, the challenges to usability they create and
+the effectiveness of standard policies and procedures
+frequently employed around password use.
+
+%Linux and Mac systems make use of the Pluggable Authentication
+%Mechanism (PAM), which weakens the assumption that authentication
+%must be done with a password (password authenication is just one
+%module available for PAM). Corporate systems typically make
+%use of centralised user directories (e.g. Active Directory, OpenLDAP),
+%which allows users to use common credentials across multiple
+%IT systems (reducing the need for users to memorise multiple
+%sets of credentials).
+
+\section{Passwords and Usability}
+
+\cite{adams1999users} famously asserted 15 years ago that security
+departments take an inappropriate approach to security risk
+by treating users as risk themselves to be controlled. They
+argue further that security compromises are either knowingly
+or unknowingly due to the way security is implemented more than
+it is the direct fault of users themselves. Their recommendation
+is that any security approaches need to be designed in
+a user-centric way not just to increase usability, but to
+reduce the likelihood of compromises ultimately.
+
+Their work is still widely cited 15 years later, but
+today, employees are increasingly using passwords in their
+workplace in addition to a pleathora of passwords for
+other services used personally. A user-centric design
+approach to a modern, enterprise security setup needs to
+be even more sympathetic to the large burden on users
+from managing so many online accounts.
+
+\cite{herley2009so} argues that users rationally reject costs
+placed upon them from security policies and invokes Kuhn's
+``Paradigm Shift''\parencite{kuhn1962structure} to suggest
+that the increasing burden placed on users with every attack
+that takes place and the additional complexity that comes
+therewith may hint at an imminent, yet dramatic change in
+how users are asked to authenticate. It is yet to be seen
+if Kuhn's observations of sudden shifts of thinking
+in the scientific community apply to a more general
+userbase such as employees or indeed the general public, but
+a strong need for true change can be taken from the analogy
+at least.
+
+So, what is the nature of the burden experienced by users?
+In the next few sections, we look at various aspects
+of password security, how users are normally advised
+and critically examine the effectiveness of that advice.
+
 \section{Password Strength}
-\section{Expiry}
+
+Password-protection accounts are vunerable to at least three
+classes of attack\parencite{florencio2014password}:
+
+\begin{itemize}
+  \item Malware infection or other direct compromise of the client machine (e.g. keyloggers)
+  \item Attacker learns the passwords via phishing, brute-force guessing or other means
+  \item Access to a particular system is gained without the actual password (e.g session hijacking, cross-site request forgery)
+\end{itemize}
+
+The notion of \emph{password strength} describes how likely
+a password is to suffer the attacker learning the password via brute-force
+guessing. A brute-force attack can be said to be \emph{online} (where the
+attacker is sending multiple login attempts directly to a system protecting
+by authentication) or \emph{offline} (where an attacker has a local copy of
+an encrypted file or hashed password, to which she can perform an unlimited
+number of attempts at cracking the password without arousing suspicion).
+
+Information Theory and entropy can allows us to gauge a password's strength
+in terms of \emph{bits}, which gives us an estimate of the likelihood of
+an attacker guessing the password in one attempt. Passwords which are
+more unlikely to guess have more bits and are thus consider stronger passwords.
+
+The advice for creating strong passwords normally takes the form of tips
+such as:
+
+\begin{itemize}
+  \item use longer passwords;
+  \item avoid using real words;
+  \item use a mix of lower and upper case; and
+  \item include numbers and other non-alphanumeric characters.
+\end{itemize}
+
+\section{Password Expiry}
 

### 2014-12-01 paper.tex
More work on chapter 2.

@@
 The notion of \emph{password strength} describes how likely
-a password is to suffer the attacker learning the password via brute-force
+a password is to suffer from an attacker learning the password via brute-force
 guessing. A brute-force attack can be said to be \emph{online} (where the
@@
 
+\cite{kaufman2002network} facetiously claimed that humans are poor
+at memorising strong passwords/keys:
+
+\begin{quotation}
+  \emph{
+  Humans are incapable of securely storing high-quality
+  cryptographic keys, and they have unacceptable speed and
+  accuracy when performing cryptographic operations. (They are
+  also large, expensive to maintain, difficult to manage, and they
+  pollute the environment. It is astonishing that these devices
+  continue to be manufactured and deployed. But they are
+  sufficiently pervasive that we must design our protocols around
+  their limitations.)
+  }
+\end{quotation}
+
+This presents a problem if users are poor at working with the
+very kind of passwords needed to avoid brute-force cracking. It
+would suggest that a user-centric design would lead to something
+more in line with abilities humans \emph{do} have.
+
 \section{Password Expiry}

### 2014-12-01 paper.tex
Written a lot more for chapter 2. Word count is getting high, so I might want to summarise soon.

@@
 
+%TODO: something about the large list of password rules
+
+Additionally, SOMEONE explored economic arguments and models
+to argue that the incredibly low, \emph{a priori} likelihood
+of a user's account being compromised leads to a monetary
+cost value far lower than that needed to justify a user
+spending any real time at all strengthening their password.
+This may be a case for a general, public account, but perhaps
+some accounts with, e.g. corporate webmail, could be classed
+as higher risk and bigger targets for attackers. An attack
+on a corporate account might allow access to systems that
+can lead to real economic or reputational damage.
+
+We can argue that these costs -- even if higher than those
+in SOMEONE's models -- are mainly suffered by the \emph{organisation}
+more than the employees. This is very similar to SOMEONE's
+arguments that even in the case of banks, they are likely
+to take on the costs of fraud and prevent their customers
+from exposure to those costs themselves.
+
+This effect is classed
+by SOMEONE as a negative externality on computer users in
+that they are taking on additional effort to cover costs
+that are not their own. These economic arguments can be
+used to make a case for organisations taking more
+responsibility and investing more to prevent the costs
+from a security compromise. This can be combined with
+the arguments for a more user-centric design
+\parencite{adams1999users} to build a case for a new
+authentication paradigm that is low friction to users and
+where economic incentives lead to an equilibrium where
+organisations spend exactly as much on security as
+is rationally justified by their real risks and costs.
+
 So, what is the nature of the burden experienced by users?
@@
 
-Password-protection accounts are vunerable to at least three
+Password-protected accounts are vunerable to at least three
 classes of attack\parencite{florencio2014password}:
@@
 
-Information Theory and entropy can allows us to gauge a password's strength
+Information Theory and entropy can allow us to gauge a password's strength
 in terms of \emph{bits}, which gives us an estimate of the likelihood of
@@
 
+Futhermore, many have noted that strong passwords defend
+against \emph{only} that single attack vector. Learning the
+password within guessing (e.g. via phishing) will put strong
+passwords in the hands of attackers just as easily as with
+weak passwords. Also, different classes of attacks entirely
+such as installing keyloggers via malware or hijacking browser
+sessions are unaffected by password strength.
+
+%TODO: who?
+Some have thus questioned whether the benefits from strong
+passwords are as substantial as standard security advice
+would suggest.
+
 \section{Password Expiry}
+
+\section{Password Reuse}
 

### 2014-12-04 paper.tex
Some rewordings.

@@
 a user-centric way not just to increase usability, but to
-reduce the likelihood of compromises ultimately.
+reduce the likelihood of compromises as well.
 
@@
 today, employees are increasingly using passwords in their
-workplace in addition to a pleathora of passwords for
-other services used personally. A user-centric design
+workplace in addition to a plethora of passwords for
+other services used personally outside of work.
+A user-centric design
 approach to a modern, enterprise security setup needs to
@@
 \cite{herley2009so} argues that users rationally reject costs
-placed upon them from security policies and invokes Kuhn's
+placed upon them from security policies. He invokes Kuhn's
 ``Paradigm Shift''\parencite{kuhn1962structure} to suggest
@@
 in the scientific community apply to a more general
-userbase such as employees or indeed the general public, but
+user base such as employees or indeed the general public, but
 a strong need for true change can be taken from the analogy
@@
 
+% TODO: many = who?
 Futhermore, many have noted that strong passwords defend
@@
 
+It should also be noted that password strength is
+used mainly to prevent
+\emph{offline} brute force attacks as many systems
+implement a policy of locking an account (or otherwise
+delaying password attempts) when too many incorrect
+password guesses are made. This effectively negates any
+\emph{online} brute force attack against even a very weak
+password.
+
 %TODO: who?
@@
 passwords are as substantial as standard security advice
-would suggest.
+would suggest. However, this is perhaps still a concern
+for an organisation as it likely does not want to suffer
+a breach in an account that had a weak password. Some
+online brute force attacks have been known to take place
+where the attackers enumerate over all known user accounts
+using a set of common or weak passwords. This increases
+the probability that they gain access to at least one
+account, which may be more useful than targeting a single
+account.
+
+So, if there is still an incentive for an organisation
+to ensure its employees' accounts are secure, but there
+are few pressures (or even rational justification)
+for the individual users to spend their time and effort doing
+so, what can the organisation do?
+
+One classical solution to weak passwords
+has been to enforce password strength \emph{requirements}.
+This is a technical solution whereby the systems simply
+reject self-selected user passwords that do not conform
+to a target strength or entropy.
+This is a technical solution that shifts the burden forcibly onto
+users and we can argue it is susceptible to the deliberate workarounds
+that many users ultimately employ to reduce the overheads
+in performing their primary jobs.
+
+Another, softer approach is to attempt to educate users
+on choosing strong, yet memorable, passwords. This belief
+that weak passwords are always down to lack of training
+is fundementally challenged in the influential findings
+of \cite{adams1999users}. Their work combined with the
+economics-based arguments and models introduced by %WHO?
+builds a large weight of evidence that well-informed
+users will still not follow much security advice
+(sometimes rationally so) to reduce their own costs in
+time and effort or to remove a significant barrier
+to performing their work tasks.
 
@@
 \section{Password Reuse}
+
+\section{The Case for a Paradigm Shift}
 

### 2014-12-06 paper.tex
First pass of chaper 2 done.

@@
 \section{Password Strength}
+\label{sec:strength}
 
@@
 
+Brute-force password guessing can also be mitigated by
+users changing passwords frequently enough that attackers
+have a short window within which to guess a particular
+password.
+
+As with password strength, this is only mitigating the
+brute-force guessing attack vector as phishing, keylogging, etc.
+can still give an attacker a password for a short while -- until
+the password is next changed. As with password strength, this
+is also most effective against offline brute-force attacks (e.g.
+the attacker has acquired a list of hashed passwords) an
+account locking policy also prevents an online brute-force
+attack without the need for passwords to expire.
+
+It is a fairly common practice in many organisations to
+implement a password expiration policy whereby the
+authentication systems will require a user to change their
+password after some defined period of time.
+
+\cite{adams1999users} note that password expiry increases
+the burden on users having to memorise multiple passwords and that
+50\% of users in their study wrote passwords down, with one
+person citing a monthly password expiration policy as the cause
+of having to do so.
+
+\cite{zhang2010security} go further to demonstrate the ease of
+algorithmically predicting a user's new password when the old,
+expired password is known. They assert that password expiration
+is weak at achieving the intent behind its implementation and
+support the idea of moving away from passwords entirely.
+
 \section{Password Reuse}
 
+Security advice given to users usually includes the suggestion that users
+have a unique password for each service they use. This can
+be an unreasonable burden on users\parencite{florencio2014password}
+and is thus an unrealistic expectation for an organisation or online
+service.
+
+\cite{preibusch2010password} showed how the pressure to reuse passwords
+presents negative externalities from security-indifferent websites upon
+security-concerned websites. This presents an organisation with
+security risks stemming solely from the existence of other, less
+secure websites with which its employees have accounts.
+
+The secret nature of passwords means an organisation cannot \emph{enforce}
+the use of unique passwords, so it must either make reuse impossible or
+reduce the impact from reuse. Businesses can make reuse impossible
+by having password strength requirements that are stronger than
+services less concerned with security, with which employeess may also have
+accounts. This is susceptible to the issues in section~\ref{sec:strength}
+where this increases costs on users over and above than those they will
+suffer from a security breach.
+
+Clearly, password reuse is a coping strategy for users arising from the
+difficulty humans have with passwords as an authentication mechanism. This
+adds weight further to the case for a real shift away from passwords in
+general. We will explore this case in section~\ref{sec:case} and evaluate
+the alternatives to password use in chapter~\ref{chapter:alternatives}.
+
 \section{The Case for a Paradigm Shift}
+
+In this chapter, we have built up the following claims:
+
+\begin{enumerate}
+  \item Strong passwords reduce guessability, but do not protect against
+    many other attacks whilst adding an economically inefficient cost to users'
+    time and effort.
+  \item Password expiration policies are similar in only protecting against
+    brute-force guessing and their effectiveness is called into question when
+    users use simple mutations to choose their next password.
+  \item Organisations and websites do not exist in isolation, but instead in
+    an ecosystem where their users have passwords on other services and are likely
+    to reuse them to reduce the burden of memorising too many.
+\end{enumerate}
+
+In summary, the use of passwords is problematic for organisations and policies
+to reduce those problems place the costs on the wrong people: the users. We can
+make the case here for a shift away from passwords in general or at least
+for further research into user-centric designs that handle the costs of security
+rationally with suitable consideration given to usability.
+
+In chapter~\ref{chapter:alternatives}, we look at some alternatives to passwords
+and evaluate them.
 

### 2014-12-09 paper.tex
Lots done on chapter 3.

@@
 
+\section{What alternatives exist?}
+
+Much of the literature talks about introducting password alternatives
+as part of two-factor authentication, i.e. introducting a second factor
+alongside a password to reduce some of the insecurities of password use
+alone. It could be argued, however, that such methods may be used as
+a \emph{replacement} for password use and there would still be an increase
+in both usability and security.
+
+Single factor based on possession (``something
+you have'') is both more intuitive (people are used to carrying credit cards
+and keys safely more than they are capable of memorising random
+cryptographic secrets) and more secure (more usability reduces the chance
+people will ``work around'' the mechanisms).
+
+Ultimately, if we can evaluate the use of each alternative as if it were
+to be used as a single-factor, it should not prevent an organisation
+adopting the alternative alongside password authentication in principle. We
+will give some consideration for the ease of doing so in our evaluation. We
+would advise always doing so as part of a design that considers the users,
+models their use cases and ensures that the approach is indeed increasing
+security more than it is just moving the costs onto the users.
+
+\cite{van2011tiqr} introduced a new alternative authentication mechanism
+called \emph{tiqr} that uses QR codes and a bespoke smartphone app. In
+doing so, he also listed and evaluated a number of existing alternatives. We
+can thus make a working list of existing non-password authentication
+mechanisms:
+
+\begin{itemize}
+  \item OTP Apps (generating One-Time Password codes with a smartphone app, e.g.
+    using HOTP or TOTP as are used by Google Authenticator)
+  \item OTP Tokens (physical tokens such as SecurID that generate one-time
+    password codes at the press of a button)
+  \item SMS OTP (sending a user one-time password codes over SMS)
+  \item PKI Tokens (a physical item such as a smartcard that contains
+    a cryptographic private/public key pair)
+  \item Mobile PKI (an approach similar to PKI tokens, but uses the SIM of a
+    mobile as the key, which allows key authentication over SMS)
+  \item Challenge/response tokens (similar to OTP tokens, but uses symmetric
+    encryption to respond to challenges)
+  \item TIQR (a web authenication approach that involves QR codes and a
+    bespoke smartphone app)
+  \item PAM USB (a module for PAM that turns any USB mass storage device
+    into a physical key to unlock or lock a computer)
+\end{itemize}
+
+PAM USB\parentcite{steele2005paranoid} is an additional mechanism added to
+this list. It allows users of Unix systems to log in by plugging in a USB
+mass storage device.
+
+In this paper, we will explore the feasibiity of adopting some of
+these authentication methods within an enterprise. The primary consideration
+is around detailing the setup and maintenance steps and whether any
+method can be justifiably ruled out in the early phases of a feasibility
+study. In section~\ref{sec:evaluation}, we will look at how the evaluation
+was carried out and the remaining sections detail that evaluation against each
+considered method.
+
+\section{Evaluating Use in the Enterprise}
+\label{sec:evaluation}
+
+Some of the identified alternatives to password authentication were
+evaluated for use in an organisational setting. The focus of this evaluation
+was to detail the steps involved in creating a toy network of machines
+that allowed common authentication across a representative sample of
+classes of services that may be used by businesses.
+
+The requirements for our toy/proof-of-concept systems were:
+
+\begin{itemize}
+  \item users can sign into desktop machines using the authentication method
+    being evaluated;
+  \item users can then access their corporate email without needing additionally
+    supplying a password to do so;
+  \item users are able to sign in to a dummy web service (a simple static page
+    behind an authentication layer was used);
+  \item services should be able to acquire information about a user for
+    authorisation purposes (e.g. determining a user's department from
+    a corporate user directory service); and
+  \item users should be able to do remote SSH login to development Unix
+    servers (this assumes a company where this is a common work task, e.g.
+    software development, which is including to exercise the authentication
+    further rather than to act as a typical business use case).
+\end{itemize}
+
+This setup has many parallels with the needs for common credentials or
+Single Sign-On (SSO) within business computer systems. It is typical for
+companies to have an authoritative directory server so that users
+can at least use the same password across all company services. Some businesses
+attempt true SSO and ensure users only type their password during initial
+login and they are automatically authorised against all revelant systems
+thereafter. This is recommended practice to reduce some of the effort
+remembering multiple passwords or even typing the same password several times
+a day. Ideally, a suitable password alternative would provide similar
+convenience in minimising effort from users when allowing access to a multitude
+of systems in a day.
+
+The methods evaluated were:
+
+\begin{itemize}
+  \item OTP Apps. This is functionally equivalent to a physical OTP token,
+    so a single evaluation was done on OTP apps only with the assumption that
+    many aspects will generalise to any OTP generator. Some consideration
+    was made around SMS OTP, but this was not evaluated directly.
+  \item PKI Tokens. A smartcard-based PKI token was evaluated with the
+    assumption again this will extrapolate to any PKI token. Mobile PKI was
+    ruled out for consideration as it requires mobile operators to install
+    keys on their customers SIM cards, which has been problematic in the
+    past.
+  \item PAM USB.
+\end{itemize}
+
+TIQR was ruled out as the aims of the project to date seem to be based around
+web-based logins only, which does not fit our defined enterprise need of
+supporting desktop computer login.
+
+\section{PKI Tokens and Smartcards}
+
+\section{OTP Mobile Apps}
+
+\section{PAM USB}
+
 \chapter{Conclusions and Recommendations}

### 2014-12-09 paper.tex
Written more and added base Vagrantfile.

@@
 
-In this paper, we will explore the feasibiity of adopting some of
+In this chapter, we will explore the feasibiity of adopting some of
 these authentication methods within an enterprise. The primary consideration
@@
 web-based logins only, which does not fit our defined enterprise need of
-supporting desktop computer login.
+supporting desktop computer login. Challenge/response tokens are also
+out of scope for this paper as they do not differ too greatly from OTP
+token and our focus is primarily on evaluating a small, but diverse, set
+of password alternatives.
 

### 2014-12-11 paper.tex
Made a fair start on chapter 3 and configured Vagrant to start an OTP network.

@@
 
+\section{OTP Mobile Apps}
+
+The chosen OTP system was the OATH Toolkit and specifically its
+module for the Unix Pluggable Authentication Module (PAM) system
+\parencite{samar1996unified}.
+
+The OATH PAM module is simple to configure on a Debian GNU/Linux desktop
+computer and provides one-time password login using any OTP token or mobile
+app, such as Google Authenticator.
+
 \section{PKI Tokens and Smartcards}
-
-\section{OTP Mobile Apps}
 

### 2014-12-12 paper.tex
Fair progress on chapter 3 now.

--- 2014-12-09 paper.tex
@@
 adopting the alternative alongside password authentication in principle. We
-will give some consideration for the ease of doing so in our evaluation. We
+gave some consideration for the ease of doing so in our evaluation. We also
 would advise always doing so as part of a design that considers the users,
@@
 
-\section{PKI Tokens and Smartcards}
-
 \section{OTP Mobile Apps}
 
+The chosen OTP system was the OATH Toolkit and specifically its
+module for the Unix Pluggable Authentication Module (PAM) system
+\parencite{samar1996unified}.
+
+The OATH PAM module was simple to configure on a Debian GNU/Linux desktop
+computer and provided one-time password login using any OTP token or mobile
+app, such as Google Authenticator. Our setup was tested with both the
+command line tool from the OATH Toolkit as well as with the
+Google Authenticator Android app.
+
+The challenge here is that each user needs a unique secret which has to
+be managed on a trusted server. The secret for every employee then needs
+to be available in a configuration file on every desktop computer. This
+can be achieved with configuration management systems such as Puppet,
+Chef, Ansible, Salt, etc.
+
+Another option is to try to centralise the OTP authentication.
+Solutions such as LinOTP suggest the abiity to have a centralised
+OTP-based authentication system with a custom PAM module. Combined
+with Kerberos, true Single Sign-On should then be possible
+including for access to email and web-based applications.
+
+A centralised solution is more flexible for corporate systems
+as PAM does not provide an intrinsic method for fowarding local
+authentication on to other services. Microsoft Windows achieved
+this in the past with NTLM -- which is possible to configure
+also on Unix systems with Winbind -- but that is largely
+regarded as an insecure mechanism today.
+
+Some desktop environments (e.g. Gnome, Mac OS X, etc.) provide
+a local keychain that can store SSH keys, SSL certificates, etc.
+which are unlocked whilst the user is logged in, but inaccessible
+otherwise. This already provides a convenience desktop
+users, but it relies on the user authenticating with a password
+that is then the same password used to update the keychain
+(Mac OS X updates the keychain password when the user's main
+account password is changed). There maybe be some scope for
+further research into whether dekstop keychains can be made
+compatible with OTP authentication.
+
+There is certainly scope for further work to evaluate extensive
+use of LinOTP and Kerberos within an enterprise setting and the
+benefits they bring over relying on a local desktop keychain.
+
 \section{PAM USB}
+
+The PAM USB module for the PAM system is similar in many
+ways to the OATH OTP module in terms of setup. The module
+installs easily on Debian GNU/Linux systems and it provides
+tools to manage a single configuration file that stores a
+mapping between users and an identifier for their USB device.
+
+The module also provides a mechanism for locking computers
+when the USB key is removed, which is helpful in corporate
+environments where users are instructed to lock their desktops
+when they are unattended. This mechanism is in line with
+drivers removing their keys when they leave their car
+unattended, so this is more in line with existing user
+habits than asking users not to disclose or to strengthen
+their passwords.
+
+However, this PAM module has the same limitations as the
+OATH OTP module in that the very simple local authenication
+is not easily forwarded to other services on the network.
+
+An advantage this module has over the OATH OTP module is that
+local applications can make authentication requests
+multiple times without the user needing to be involved. As
+long as the USB device remains connected, applications
+can authenticate multiple times, but OTP authentication
+will require a new password to be entered on each authentication
+request.
+
+\section{PKI Tokens and Smartcards}
 

### 2014-12-12 paper.tex
written up on OpenPGP card

@@
   \item PAM USB.
+  \item YubiKey. This is a hardware USB device from Yubico that can act
+    as an OTP generator, a challenge-response token and even a store
+    for a static password.
 \end{itemize}
@@
 
+%An advantage this module has over the OATH OTP module is that
+%local applications can make authentication requests
+%multiple times without the user needing to be involved. As
+%long as the USB device remains connected, applications
+%can authenticate multiple times, but OTP authentication
+%will require a new password to be entered on each authentication
+%request.
+
 An advantage this module has over the OATH OTP module is that
-local applications can make authentication requests
-multiple times without the user needing to be involved. As
-long as the USB device remains connected, applications
-can authenticate multiple times, but OTP authentication
-will require a new password to be entered on each authentication
-request.
+a USB storage device can also store other authentication
+information such as SSH keys, SSL client certificates or
+anything else to aid the Single Sign-On experience. This
+increases the impact of the device being lost unless
+these private keys are secured with a password, which
+reintroduces passwords back into the user's workflow. Alternatively,
+the risks of loss could be mitigated with an effective
+key and certificate revocation process.
+
+Futher research may find a solution to storing keys
+and certificates on a USB device without passwords or highlight
+whether the risks of doing so are too high.
 
 \section{PKI Tokens and Smartcards}
+
+Our test desktop machine was configured to accepted login
+with a smartcard reader and an OpenPGP smartcard
+\parencite{pietig2009functional}. The OpenPGP card integrates
+with GnuPG to provide a public-private key pair that can
+be used for encryption.
+
+GnuPG has a gpg-agent service that allows these keys to
+be used for SSH authentication as well, which can provide
+a Single Sign-On experience betweeen logging in to a local
+desktop and connecting out to remote servers.
+
+For desktop login, we were able to use the ``Poldi'' PAM
+module. The use of the smartcard requires a PIN and
+the card has a lock-out after 3 incorrect tries. This is
+behaviour users can recognise from their bank smartcards
+and the lock-out feature negates the guessability of
+even a weak 4 to 6 digit PIN.
+
+The use of a PIN does not eliminate the ``something you
+know'' factor, but a short PIN is far less of a burden
+than the requirement for users to have strong, random
+passwords. This could be a reasonable compromise for an
+enterprise given some further usability research.
+
+\section{YubiKey}
 

### 2014-12-13 paper.tex
More work on chapter 3.

@@
 is rationally justified by their real risks and costs.
+
+\cite{beautement2009compliance} introduce the idea of a
+``compliance budget'' which models how users perceive these costs
+and benefits. Their work suggests a cap on the effectiveness of any
+security measures within an organisation, but also introduces management
+approaches to raise the threshold above which a user will no longer
+comply with security policies. This arguably increases the cost-benefit
+ratio by making it more worthwhile for users to practice secure
+behaviour (even if by increasing sanctions such that avoidance thereof becomes
+more beneficial than working around costs from complying with security
+requirements). They still admit there is a finite limit to improvements to
+this compliance budget through management, training and cultural changes.
+This supports the idea that many of the costs simply must be moved to the
+organisation itself.
 
@@
 
-Our test desktop machine was configured to accepted login
-with a smartcard reader and an OpenPGP smartcard
-\parencite{pietig2009functional}. The OpenPGP card integrates
-with GnuPG to provide a public-private key pair that can
-be used for encryption.
+Our test desktop machine was configured to accepted login with a smartcard
+reader and an OpenPGP smartcard \parencite{pietig2009functional}. The OpenPGP
+card integrates with GnuPG to provide a public-private key pair that can
+be used for encryption. For desktop login, we were able to use the ``Poldi''
+PAM module.
 
-GnuPG has a gpg-agent service that allows these keys to
-be used for SSH authentication as well, which can provide
-a Single Sign-On experience betweeen logging in to a local
-desktop and connecting out to remote servers.
+GnuPG has a gpg-agent service that allows these keys to be used for SSH
+authentication as well, which can provide a Single Sign-On experience
+between logging in to a local desktop and connecting out to remote servers.
 
-For desktop login, we were able to use the ``Poldi'' PAM
-module. The use of the smartcard requires a PIN and
-the card has a lock-out after 3 incorrect tries. This is
-behaviour users can recognise from their bank smartcards
-and the lock-out feature negates the guessability of
-even a weak 4 to 6 digit PIN.
+The use of the
+smartcard requires a PIN and the card has a lock-out after 3 incorrect tries.
+This is behaviour users can recognise from their bank smartcards and the
+lock-out feature negates the guessability of even a weak 4 to 6 digit PIN.
 
@@
 enterprise given some further usability research.
+
+The difficulty of eliminating a password or PIN from the login workflow
+suggests that the PKI token or smartcard approach is more suitable for
+situations where security requirements are higher and two-factor authentication
+is preferable. The same technology is used for debit/credit card transactions
+and for withdrawing cash from a bank account and thus we would expect it
+to be used in other organisations that have similar security requirements.
+
+The \emph{Scute} project provides a PKCS~\#11 module that allows an OpenPGP
+smartcard to be used with the Network Security Services (NSS) within
+the Firefox web browser (we believe Chrome also supports this, but Firefox
+was used in our test network). Combined with Apache's \emph{mod_nss} (which
+derives from mod_ssl), we were able to implement an authentication layer
+over a web-based application that supported an OpenPGP smartcard.
+
+NSS also allows the same key to be used within Mozilla Thunderbird for
+email authentication
 

### 2014-12-13 paper.tex
first draft of chapter 3 done

@@
 \section{OTP Mobile Apps}
+\label{sec:otp}
 
@@
 NSS also allows the same key to be used within Mozilla Thunderbird for
-email authentication
+email authentication. Alternatively, the OpenSC project provides an
+OpenSSL engine for PKCS#11 implementations with which we can store an X.509
+client certificate on the smartcard itself.
+
+Given the wide variety of
+options for integrating with services such as email, we did not go as far
+as to evaluate them all and a thorough examination of a smartcard-based,
+enterprise computer network is a subject for further research. The most
+subtantial work in this area (that we know of) is from \cite{grossesecuring}.
+This research compares different smartcard technologies and software projects
+and evaluates them for functionality that overlaps with the enterprise
+requirements we are following in this paper. The information seems to be
+quite dated now, 8 years later, as we have mentioned projects in this
+section that cover functionality claimed not to exist -- especially for OpenPGP
+cards.
 
 \section{YubiKey}
+
+The YubiKey \parencite{merkel2009yubikey} is a hardware device
+manufactured by Yubico that supports TOTP, HOTP, storing a static password and
+a challenge-response mechanism. Yubico provide a substantial amount of
+open source tooling and form enterprise partnerships to support a wide
+range of applications.
+
+In our test network, we were able to make the YubiKey serve the functionality
+of \texttt{pam\_oath} by using as an OTP token. This opens up the possibility
+of providing a YubiKey to employees, but also allowing use of mobile apps
+such as Google Authenticator if the key is not available (or simply increase
+usability by giving users the choice). This might be a crucial option to
+allow users to login in to a self-service application with which they can
+report their YubiKey as lost/stolen and de-associate it with their user
+account.
+
+This also means all the benefits of OTP tokens as outlined in
+section~\ref{sec:otp} apply also to the YubiKey. An additional benefit,
+however, is that the YubiKey has a button on the device that types the
+one-time password for the user, which allows not only for more secure tokens,
+but also increases the overhead of the login process. The YubiKey achieves
+this by acting as a USB keyboard during this operation.
+
+Further to this, Yubico provide their own PAM module that has a challenge-response
+mode that can automatically detect if the user's YubiKey is plugged in
+and log the user in without any interaction. This means the YubiKey provides
+the functionality of \texttt{pam\_usb} as well.
+
+A drawback is that the YubiKey lacks additional storage for other certificates
+or keys (as is the case with a USB storage device or an OpenPGP smartcard). This
+makes it difficult to forward the authentication to remote systems. Yubico
+produce the more expensive YubiKey Neo which supports OpenPGP and acts as
+a smartcard. We did not evaluate this model in our test network.
+
+Perhaps the most promising feature of the YubiKey for business use is the
+ability to store a static password as well as the functionality above. A
+managed distrubution of YubiKey hardward to employees within an organisation
+could involve setting up each key with a strong, static password. The users
+would not need to memorise the password, but it could be used as part of
+a shared-credentials setup (e.g. LDAP directory service already established
+within the organisation) without any significant changes to the technology
+used.
+
+In chapter~\ref{chapter:conclusion}, we explore the implication of the
+static password feature within a company and draw some conclusions about the
+different password alternatives and their use in the enterprise.
 

### 2014-12-13 paper.tex
First pass done. Needs lots of cutting.

@@
 \begin{itemize}
-  \item users can sign into desktop machines using the authentication method
+  \item \textbf{Desktop Login}: users can sign into desktop machines using the authentication method
     being evaluated;
-  \item users can then access their corporate email without needing additionally
+  \item \textbf{Company Email}: users can then access their corporate email without needing additionally
     supplying a password to do so;
-  \item users are able to sign in to a dummy web service (a simple static page
+  \item \textbf{Web Login}: users are able to sign in to a dummy web service (a simple static page
     behind an authentication layer was used);
-  \item services should be able to acquire information about a user for
-    authorisation purposes (e.g. determining a user's department from
-    a corporate user directory service); and
-  \item users should be able to do remote SSH login to development Unix
+  \item \textbf{SSH Login}: users should be able to do remote SSH login to development Unix
     servers (this assumes a company where this is a common work task, e.g.
@@
 
-Some desktop environments (e.g. Gnome, Mac OS X, etc.) provide
+LinOTP also provides centralised web-based identity approaches
+such as OpenID and SAML. A company-wide identity provider would
+not only allow Single Sign-On within a series of web-based
+internal applications, but opens up the possibility of employees
+being able to use their company identity elsewhere. One example
+is if the company wishes to purchase access to online resources
+(online books, journals or any other cloud-based service), they
+could make agreement with the provider to allow employees to log in
+via OpenID or SAML. Universities already use this model for access to online
+journals and academic materials.
+
+Some desktop environments (e.g. Gnome, Mac OS X) provide
 a local keychain that can store SSH keys, SSL certificates, etc.
@@
 
+An Apache module is also available
+for both the Yubikey and also generic OTP logins. Whilst having multiple
+login points all asking for a new OTP would be an inconvenience when using a mobile
+app, this is less of a problem for the YubiKey where the user needs only press
+a single button each time. This might allow for internal applications each to
+implement their own OTP support and for them to be managed separately, whereas
+the OTP mobile app approach would be better with true Single Sign-On.
+
 A drawback is that the YubiKey lacks additional storage for other certificates
@@
 
+Table~\ref{table:overview} summarises the approaches tried on our test network
+and discussed in chapter~\ref{chapter:alternatives}. In some cases, there
+are a few alternatives listed where some aspects of more than one approach
+were explored. These are by no means a definitive list of all options available
+to individuals and organisations and some of the technologies and systems
+not evaluated thoroughly are mentioned in passing throughout
+chapter~\ref{chapter:alternatives}.
+
+\begin{table}[h]
+  \begin{tabular}{|l|l|l|l|l|} \hline
+                            & \textbf{OTP Apps} & \textbf{pam\_usb} & \textbf{OpenPGP smartcard} & \textbf{YubiKey} \\ \hline
+    \textbf{Desktop Login}  & pam\_oath        & pam\_usb                    & pam\_poldi & pam\_oath, pam\_yubico  \\ \hline
+    \textbf{Company Email}  & LinOTP + Kerberos & X.509 certificate on device & Scute      & LinOTP + Kerberos, static password + OpenLDAP   \\ \hline
+    \textbf{Web Login}      & OpenID, SAML     & X.509 certificate on device & Scute      & Apache YubiKey/OTP modules, static password + OpenLDAP  \\ \hline
+    \textbf{SSH Login}      & Gnome keyring    & SSH key stored on device    & gpg-agent  &  \\ \hline
+  \end{tabular}
+  \label{table:overview}
+\end{table}
+
+Overall, YubiKeys seem to provide the best balance between application support,
+usability and compatibility. OpenPGP and other smartcards are perhaps also
+a good approach where security is of greater concern and there is some appeal
+in terms of usability given users are already used to keeping credit cards
+safe and protecting them with short PINs. Their requirement for specialised
+card readers and the fact it is still best practice to protect them with memorised
+information means significant monetary and effort costs that may be acceptable
+to trade off for the additional security, but this must be considered.
+
+LinOTP is a promising enterprise solution that is compatible with Windows
+domain setups common in a lot of companies. it appears to provide the most
+flexibility in terms of application support.
+
+Perhaps the most important YubiKey feature in practical terms is the ability
+to store a static password alongside its OTP behaviour. This enables its
+use in \emph{existing} corporate networks where users have a single username-password
+pair and all systems authenticate against a directory service such as Active Directory
+or OpenLDAP. To set this up, security departments need only include synchronsing
+users' passwords to their YubiKeys as part of account creation. Users' passwords
+can thus become much stronger and more random without the risk of forgetting or
+users resorting to writing them down.
+
+The static password feature is entirely compatible with every system that already
+supports password authentication, so the only additional costs are purchasing
+the hardware and ensuring users are given quick replacements with new passwords
+when they report their existing key lost or stolen. Users are more aware of the
+need to report lost or stolen physical items (company ID card, house keys, credit cards, etc.)
+whereas the evidence in the literature around passwords suggests users disclose
+passwords far more readily (whether deliberately or inadvertantly).
+
+Thus we can recommend coporations are able to evaluate the hardware costs
+and consider YubiKeys as a ``drop in'' replacement for users memorising
+passwords with few infrastructure changes. This strategy even lends itself
+to the possiblity of being run as a trial with one team or department. Over time,
+more structural changes such as introducing LinOTP could be considered to
+reap some of the benefits OTP and challenge-response bring (e.g. protection
+from replay attacks, reduced exposure to brute-force guessing).
+
+It may be possible to run the static password trial alongside introducing OTP
+for some applications only. The YubiKey supports generating either token
+by using a short versus a long button press, but the usability of that has
+not been demonstrated.
+
 \chapter{Further Work}
+
+The following questions are open and are possibilities for further research:
+
+\begin{itemize}
+\item What is the effectiveness and what potential issues are there around
+  a real deployment of LinOTP within a business? An empirical study in a
+  real world setting would give some valuable insights.
+\item What is the best approach (if any) to storing keys and certificates on
+  a USB device, keeping them secure and allowing them to be used in onward,
+  automatic authentication against other services after login? Can a user's
+  own USB storage device (already commonplace in many offices) be thus set
+  up as an authentication token without the need for specialist hardware
+  such as YubiKeys?
+\item Today, how easily can an IT or security department set up users with
+  smartcards to provide a Single Sign-On experience across all company
+  systems? A real world case study would be really useful to bring together
+  all the individual projects' documentation and prove an end-to-end deployment
+  of them.
+\item Are users happy to treat smartcards such as OpenPGP cards within the
+  same guidelines already followed for credit cards?
+  \item What real world case studies exist for enterprise deployments of YubiKeys?
+    Yubico claim multiple corporate customers on their website, but have any
+    been studied in the academic literature?
+\item Many of the mechanisms discussed are seeing large-scale
+  adoption as a second factor in addition to passwords. However, What are the
+  security implications of relying on any password alternative as a single factor?
+  Do any of them end up being more vulnerable than a traditional, memorised
+  password when used in isolation?
+\end{itemize}
 

### 2014-12-13 paper.tex
Got LaTeX build to pass.

@@
 \section{The Case for a Paradigm Shift}
+\label{sec:case}
 
@@
 
-PAM USB\parentcite{steele2005paranoid} is an additional mechanism added to
+PAM USB\parencite{steele2005paranoid} is an additional mechanism added to
 this list. It allows users of Unix systems to log in by plugging in a USB
@@
 the Firefox web browser (we believe Chrome also supports this, but Firefox
-was used in our test network). Combined with Apache's \emph{mod_nss} (which
-derives from mod_ssl), we were able to implement an authentication layer
+was used in our test network). Combined with Apache's \emph{mod\_nss} (which
+derives from mod\_ssl), we were able to implement an authentication layer
 over a web-based application that supported an OpenPGP smartcard.
@@
 email authentication. Alternatively, the OpenSC project provides an
-OpenSSL engine for PKCS#11 implementations with which we can store an X.509
+OpenSSL engine for PKCS\#11 implementations with which we can store an X.509
 client certificate on the smartcard itself.
@@
 \chapter{Conclusions and Recommendations}
+\label{chapter:conclusion}
 

### 2014-12-13 paper.tex
Starting to cut the word count down.

@@
 The use of passwords for authentication is certainly commonplace.
-% TODO: How many passwords?
 It is standard practice within small and large enterprises alike
@@
 
-%TODO: something about the large list of password rules
-
 Additionally, SOMEONE explored economic arguments and models
@@
 
-Information Theory and entropy can allow us to gauge a password's strength
-in terms of \emph{bits}, which gives us an estimate of the likelihood of
-an attacker guessing the password in one attempt. Passwords which are
-more unlikely to guess have more bits and are thus consider stronger passwords.
-
 The advice for creating strong passwords normally takes the form of tips
@@
 
-\cite{kaufman2002network} facetiously claimed that humans are poor
-at memorising strong passwords/keys:
-
-\begin{quotation}
-  \emph{
-  Humans are incapable of securely storing high-quality
-  cryptographic keys, and they have unacceptable speed and
-  accuracy when performing cryptographic operations. (They are
-  also large, expensive to maintain, difficult to manage, and they
-  pollute the environment. It is astonishing that these devices
-  continue to be manufactured and deployed. But they are
-  sufficiently pervasive that we must design our protocols around
-  their limitations.)
-  }
-\end{quotation}
-
 This presents a problem if users are poor at working with the
@@
 
-% TODO: many = who?
-Futhermore, many have noted that strong passwords defend
-against \emph{only} that single attack vector. Learning the
-password within guessing (e.g. via phishing) will put strong
-passwords in the hands of attackers just as easily as with
-weak passwords. Also, different classes of attacks entirely
-such as installing keyloggers via malware or hijacking browser
-sessions are unaffected by password strength.
+\cite{florencio2007large} noted that strong passwords are just
+as easily stolen through phishing and key logging as weak
+passwords. Password strength is thus limited in only protecting
+brute force attacks and not problems where users accidentally
+or deliberately disclose their password.
 
@@
 
-%TODO: who?
-Some have thus questioned whether the benefits from strong
-passwords are as substantial as standard security advice
-would suggest. However, this is perhaps still a concern
-for an organisation as it likely does not want to suffer
+The real concern for an organisation
+is that it likely does not want to suffer
 a breach in an account that had a weak password. Some
@@
 where the attackers enumerate over all known user accounts
-using a set of common or weak passwords. This increases
+using a set of common or weak passwords -- sometimes known
+as a \emph{reverse brute force} attack \parencite{khan2010comparative}.
+This increases
 the probability that they gain access to at least one

### 2014-12-13 paper.tex
Fixed citations.

@@
 
-Their work is still widely cited 15 years later, but
-today, employees are increasingly using passwords in their
-workplace in addition to a plethora of passwords for
-other services used personally outside of work.
-A user-centric design
-approach to a modern, enterprise security setup needs to
-be even more sympathetic to the large burden on users
-from managing so many online accounts.
-
 \cite{herley2009so} argues that users rationally reject costs
@@
 
-Additionally, SOMEONE explored economic arguments and models
+Additionally, Herley explored economic arguments
+and models
 to argue that the incredibly low, \emph{a priori} likelihood
@@
 
-We can argue that these costs -- even if higher than those
-in SOMEONE's models -- are mainly suffered by the \emph{organisation}
-more than the employees. This is very similar to SOMEONE's
+We can argue that these costs
+are mainly suffered by the \emph{organisation}
+more than the employees. This is very similar Herley's
 arguments that even in the case of banks, they are likely
@@
 This effect is classed
-by SOMEONE as a negative externality on computer users in
+by Herley as a negative externality on computer users in
 that they are taking on additional effort to cover costs
@@
 
-\cite{beautement2009compliance} introduce the idea of a
+\cite{beautement2009compliance} introduced the idea of a
 ``compliance budget'' which models how users perceive these costs
@@
 Password-protected accounts are vunerable to at least three
-classes of attack\parencite{florencio2014password}:
+classes of attack \parencite{florencio2014password}:
 
@@
 
-\cite{florencio2007large} noted that strong passwords are just
+\textcite{florencio2007large} noted that strong passwords are just
 as easily stolen through phishing and key logging as weak
@@
 is fundementally challenged in the influential findings
-of \cite{adams1999users}. Their work combined with the
+of \textcite{adams1999users}. Their work combined with the
 economics-based arguments and models introduced by %WHO?
@@
 
-\cite{adams1999users} note that password expiry increases
+\textcite{adams1999users} note that password expiry increases
 the burden on users having to memorise multiple passwords and that
@@
 
-\cite{zhang2010security} go further to demonstrate the ease of
+\textcite{zhang2010security} go further to demonstrate the ease of
 algorithmically predicting a user's new password when the old,
@@
 
-\cite{preibusch2010password} showed how the pressure to reuse passwords
+\textcite{preibusch2010password} showed how the pressure to reuse passwords
 presents negative externalities from security-indifferent websites upon
@@
 
-\cite{van2011tiqr} introduced a new alternative authentication mechanism
+\textcite{van2011tiqr} introduced a new alternative authentication mechanism
 called \emph{tiqr} that uses QR codes and a bespoke smartphone app. In
@@
 enterprise computer network is a subject for further research. The most
-subtantial work in this area (that we know of) is from \cite{grossesecuring}.
+subtantial work in this area (that we know of) is from \textcite{grossesecuring}.
 This research compares different smartcard technologies and software projects

### 2014-12-14 paper.tex
cut down intro a bit

@@
 in terms of usability for users and in financial terms for the organisation
-that must implement any security processes and policies. However,
-whilst many of the benefits and requirements listed above should justify
-these costs, they are notably gains (or prevention of losses) for
-the organisation itself more than they are gains for the individual
-employees.
+that must implement any security processes and policies. In this paper, we
+note that these benefits are primarily for the organisation more than the
+users, so are the usability and effort costs placed on users justifiable
+if they are to gain very little?
 
@@
 
-User authentication that requires two differing factors
-of authentication (known as \emph{two-factor authentication}. There
+There
 has been a rise in two-factor authenication with companies
@@
 The mainstream adoption of authentication methods other than
-passwords suggests increased ease in enterprises adoption means
-of verifying users either instead of or in addition to
+passwords suggests organisations may find it easier than ever to adopt
+means of verifying users either instead of or in addition to
 existing password setups.

### 2014-12-14 logbook.tex
Starting writing up GPG log.

\documentclass{report}

\usepackage[style=authoryear,backend=biber]{biblatex}

\title{The Folly of Passwords}
\author{Ross Fenning}

\bibliography{references}

\begin{document}

\maketitle
\tableofcontents

\chapter{Wireshark Analysis}

\chapter{GNU Privacy Guard (GPG)}

In this report, we look at how to set up GPG on a Debian or Ubuntu GNU/Linux
system although many of the steps apply to other Linux and Unix systems.

The steps outlined below will demonstrate how to:

\begin{enumerate}
\item generate a new private/public GPG key pair;
\item use the public key to encrypt a plain text message;
\item use the private key to descrypt the message; and
\item have a someone sign your new to verify that you own it.
\end{enumerate}

\section{Creating a GPG key pair}

The gpg tool is normally pre-installed in a lot of Linux/Unix systems. On
a Debian or Ubuntu GNU/Linux system that does not have it installed, it
can be installed quite simply with:

\begin{listing}
\$ sudo aptitude install gpg
\end{listing}

We can run the command to see there are no keys already generated:

\begin{listing}
  \$ gpg --list-keys
  gpg: directory `/home/ross/.gnupg' created
  gpg: new configuration file `/home/ross/.gnupg/gpg.conf' created
  gpg: WARNING: options in `/home/ross/.gnupg/gpg.conf' are not yet active during this run
  gpg: keyring `/home/ross/.gnupg/pubring.gpg' created
  gpg: /home/ross/.gnupg/trustdb.gpg: trustdb created
\end{listing}

Note that on the first use of the command, it creates some configuration files
and keyrings. If there are any keys already stored in the user's home
directory, then details of them would be printed out as well.

To create a key, we simply need to run:

\begin{listing}
$ gpg --gen-key
gpg (GnuPG) 1.4.12; Copyright (C) 2012 Free Software Foundation, Inc.
This is free software: you are free to change and redistribute it.
There is NO WARRANTY, to the extent permitted by law.

Please select what kind of key you want:
   (1) RSA and RSA (default)
   (2) DSA and Elgamal
   (3) DSA (sign only)
   (4) RSA (sign only)
Your selection? 1
RSA keys may be between 1024 and 4096 bits long.
What keysize do you want? (2048) 2048
Requested keysize is 2048 bits
Please specify how long the key should be valid.
         0 = key does not expire
      <n>  = key expires in n days
      <n>w = key expires in n weeks
      <n>m = key expires in n months
      <n>y = key expires in n years
Key is valid for? (0) 0
Key does not expire at all
Is this correct? (y/N) y

You need a user ID to identify your key; the software constructs the user ID
from the Real Name, Comment and Email Address in this form:
    "Heinrich Heine (Der Dichter) <heinrichh@duesseldorf.de>"

Real name: Ross Fenning
Email address: gpg@rossfenning.co.uk
Comment:
You selected this USER-ID:
    "Ross Fenning <gpg@rossfenning.co.uk>"

Change (N)ame, (C)omment, (E)mail or (O)kay/(Q)uit? O
You need a Passphrase to protect your secret key.

We need to generate a lot of random bytes. It is a good idea to perform
some other action (type on the keyboard, move the mouse, utilize the
disks) during the prime generation; this gives the random number
generator a better chance to gain enough entropy.
\end{listing}

\section{Encrypting a message}

\section{Decrypting the message}

\section{Signing the key}

\chapter{Ethical Hacking Lab Setup}

\end{document}

### 2014-12-14 logbook.tex
Finished report on using GPG.

@@
 
+\usepackage{listings}
 \usepackage[style=authoryear,backend=biber]{biblatex}
+
+\lstset{ %
+  basicstyle=\footnotesize,
+  breakatwhitespace=true,
+  breaklines=true,
+  extendedchars=true,
+  frame=single,
+  showspaces=false,
+  showstringspaces=false
+}
 
@@
 
-\chapter{GNU Privacy Guard (GPG)}
+\chapter{GNU Privacy Guard}
 
-In this report, we look at how to set up GPG on a Debian or Ubuntu GNU/Linux
+In this report, we look at how to set up GNU Privacy Guard (GPG)
+on a Debian or Ubuntu GNU/Linux
 system although many of the steps apply to other Linux and Unix systems.
@@
 \item generate a new private/public GPG key pair;
-\item use the public key to encrypt a plain text message;
-\item use the private key to descrypt the message; and
-\item have a someone sign your new to verify that you own it.
+\item have a someone sign your new public key to verify that you own it;
+\item use the public key to encrypt a plain text message; and
+\item use the private key to descrypt the message.
 \end{enumerate}
@@
 
-The gpg tool is normally pre-installed in a lot of Linux/Unix systems. On
+The gpg tool is normally pre-installed in a lot of Linux/Unix systems as
+tools like \texttt{apt} depends on it. On
 a Debian or Ubuntu GNU/Linux system that does not have it installed, it
@@
 
-\begin{listing}
-\$ sudo aptitude install gpg
-\end{listing}
+\begin{lstlisting}[language=bash]
+$ sudo apt-get install gnupg
+[...]
+$
+\end{lstlisting}
 
@@
 
-\begin{listing}
-  \$ gpg --list-keys
+\begin{lstlisting}[language=bash]
+$ gpg --list-keys
 gpg: directory `/home/ross/.gnupg' created
@@
 gpg: /home/ross/.gnupg/trustdb.gpg: trustdb created
-\end{listing}
+$
+\end{lstlisting}
 
@@
 
-\begin{listing}
+\begin{lstlisting}[language=bash]
 $ gpg --gen-key
@@
 You need a Passphrase to protect your secret key.
+[...]
+gpg: key B8D85C76 marked as ultimately trusted
+public and secret key created and signed.
 
-We need to generate a lot of random bytes. It is a good idea to perform
-some other action (type on the keyboard, move the mouse, utilize the
-disks) during the prime generation; this gives the random number
-generator a better chance to gain enough entropy.
-\end{listing}
+gpg: checking the trustdb
+gpg: 3 marginal(s) needed, 1 complete(s) needed, PGP trust model
+gpg: depth: 0  valid:   1  signed:   0  trust: 0-, 0q, 0n, 0m, 0f, 1u
+pub   4096R/B8D85C76 2014-12-14
+      Key fingerprint = BDFD 1EF9 9052 4326 BA89  3897 30BC 9067 B8D8 5C76
+uid                  Ross Fenning <gpg@rossfenning.co.uk>
+sub   4096R/CEEFF134 2014-12-14
+$
+\end{lstlisting}
+
+After answering all the questions and providing a passphrase,
+you may be asked to generate some entropy.
+Machines had a hard time using a good source of randomness to make sure
+the key is not predictable and guessable, so you may be required to do
+some work via keystrokes, mouse movements or run some processes that
+do disk activity. Once this is complete, you will have your public and
+private key pair.
+
+To see the public key, enter:
+
+\begin{lstlisting}[language=bash]
+$ gpg --list-keys
+/home/ross/.gnupg/pubring.gpg
+-----------------------------
+pub   4096R/B8D85C76 2014-12-14
+uid                  Ross Fenning <gpg@rossfenning.co.uk>
+sub   4096R/CEEFF134 2014-12-14
+$
+\end{lstlisting}
+
+\noindent and to see the corresponding private key:
+
+\begin{lstlisting}[language=bash]
+$ gpg --list-secret-keys
+/home/ross/.gnupg/secring.gpg
+-----------------------------
+sec   4096R/B8D85C76 2014-12-14
+uid                  Ross Fenning <gpg@rossfenning.co.uk>
+ssb   4096R/CEEFF134 2014-12-14
+$
+\end{lstlisting}
+
+Note that the IDs (in our case B8D85C76) match on both the public
+and private keys, which indicates they are a complementing pair.
+
+\section{Distributing the public key}
+
+After we have successfully created our key pair, we need to send
+out our public key to people we know so that they may send
+encrypted files or messages that only we can decrypt.
+
+To export a public key into a file, do the following:
+
+\begin{lstlisting}[language=bash]
+$ gpg --output mypublickey.gpg --export B8D85C76
+$
+\end{lstlisting}
+
+This creates a file called \texttt{mypublickey.gpg} which contains
+a binary format of the key. This is suitable for sending as a file
+(e.g. SCP, FTP or email attachment). Sometimes it may be more
+convenient to export the key in ASCII for ease of pasting into emails
+or instant messaging chats. To do this, we need to \emph{ASCII-armor}
+the key thus:
+
+\begin{lstlisting}[language=bash]
+$ gpg --armor --export B8D85C76
+-----BEGIN PGP PUBLIC KEY BLOCK-----
+Version: GnuPG v1.4.12 (GNU/Linux)
+
+[...]
+-----END PGP PUBLIC KEY BLOCK-----
+$
+\end{lstlisting}
+
+The output from the previous command can be copied and pasted anywhere
+you wish your key to be available. Either the binary form or the
+ASCII-armored form of the key are safe to make available publicly for
+anyone who wishes to send you encrypted messages.
+
+Another party can then simply import your key:
+
+\begin{lstlisting}[language=bash]
+$ gpg --import mypublickey.pgp
+gpg: key B8D85C76: public key "Ross Fenning <gpg@rossfenning.co.uk>" imported
+gpg: Total number processed: 1
+gpg:               imported: 1  (RSA: 1)
+$
+\end{lstlisting}
+
+Once someone has your public key, they should verify it to ensure
+it wasn't tampered and replaced in transit. To do this, the other
+party would check your public key's fingerprint:
+
+\begin{lstlisting}[language=bash]
+$ gpg --fingerprint B8D85C76
+pub   4096R/B8D85C76 2014-12-14
+      Key fingerprint = BDFD 1EF9 9052 4326 BA89  3897 30BC 9067 B8D8 5C76
+uid                  Ross Fenning <gpg@rossfenning.co.uk>
+sub   4096R/CEEFF134 2014-12-14
+
+$
+
+\end{lstlisting}
+
+The other person would then ask you to run the same command and confirm
+over a different channel (e.g. a telephone call) that you both see
+the same fingerprint. This is arguably the most important step to get
+right as the security GPG brings is invalidated entirely if you
+trust a key not genuinely created by it's reported owner.
+
+Now the other party is happy they have imported your public key and
+verified it is indeed your key in their possession, they are then
+able to sign it:
+
+\begin{lstlisting}[language=bash]
+$ gpg --sign-key B8D85C76
+
+pub  4096R/B8D85C76  created: 2014-12-14  expires: never       usage: SC  
+                     trust: unknown       validity: unknown
+sub  4096R/CEEFF134  created: 2014-12-14  expires: never       usage: E   
+[ unknown] (1). Ross Fenning <gpg@rossfenning.co.uk>
+
+pub  4096R/B8D85C76  created: 2014-12-14  expires: never       usage: SC  
+                     trust: unknown       validity: unknown
+ Primary key fingerprint: BDFD 1EF9 9052 4326 BA89  3897 30BC 9067 B8D8 5C76
+
+     Ross Fenning <gpg@rossfenning.co.uk>
+
+Are you sure that you want to sign this key with your
+key "Other Party <other.party@example.com>" (164DFEEB)
+
+Really sign? (y/N) y
+
+You need a passphrase to unlock the secret key for
+user: "Other Party <other.party@example.com>"
+4096-bit RSA key, ID 12345678, created 2014-01-01
+
+$
+\end{lstlisting}
+
+The other party is prompted for the passphrase of \emph{their} private key
+which is then used to sign your public key within their keyring. This
+tells their system that they vouch for that key. Typically, the other
+party would then export your now-signed key again:
+
+\begin{lstlisting}[language=bash]
+$ gpg --output B8D85C76-signedby-12345678.gpg  --export B8D85C76
+$
+\end{lstlisting}
+
+\noindent so that you can then import the signed version to replace your
+unsigned one:
+
+\begin{lstlisting}[language=bash]
+$ gpg --import B8D85C76-signedby-12345678.gpg
+gpg: key B8D85C76: "Ross Fenning <gpg@rossfenning.co.uk>" 1 new signature
+gpg: Total number processed: 1
+gpg:         new signatures: 1
+gpg: 3 marginal(s) needed, 1 complete(s) needed, PGP trust model
+gpg: depth: 0  valid:   1  signed:   0  trust: 0-, 0q, 0n, 0m, 0f, 1u
+$
+\end{lstlisting}
+
+Note the ``1 new signature'' which shows your copy of your own public key
+now includes a signature by the other party verifiying that they
+believe that key is geuinely your public key. You can now export this
+version of the key to other people and they may benefit from being able
+to skip the step of verifying the fingerprint over the phone if they
+can see your key is signed by another key they already trust.
+
+This ``Web of Trust'' model is how GPG verifies identity and true ownership
+of public keys.
 
@@
 
+Now that the other party has a copy of your public key and they are certain
+that is really is your key, they can send encrypted files or messages to
+you that only you may see.
+
+For example, let us say they created a private message in a plain text file:
+
+\begin{lstlisting}[language=bash]
+$ echo "Hello" >secret.txt
+$ cat secret.txt 
+Hello
+\end{lstlisting}
+
+they are then able to encrypt this secret message so that only you may read it:
+
+\begin{lstlisting}[language=bash]
+$ echo "Hello" >secret.txt
+$ cat secret.txt 
+Hello
+\end{lstlisting}
+
+\begin{lstlisting}[language=bash]
+$ gpg --recipient B8D85C76 --encrypt secret.txt
+gpg: checking the trustdb
+gpg: 3 marginal(s) needed, 1 complete(s) needed, PGP trust model
+gpg: depth: 0  valid:   2  signed:   1  trust: 0-, 0q, 0n, 0m, 0f, 2u
+gpg: depth: 1  valid:   1  signed:   0  trust: 1-, 0q, 0n, 0m, 0f, 0u
+gpg: next trustdb check due at 2015-04-20
+$ ls secret.txt*
+secret.txt  secret.txt.gpg
+\end{lstlisting}
+
+Now we have the original message intact, but also an encrypted form of it.
+Again, the \texttt{--armor} option would have created an ASCII form of
+the encrypted message more suitable for pasting into emails (in a file
+called secret.txt.asc).
+
 \section{Decrypting the message}
 
-\section{Signing the key}
+The other party can then send the encrypted file or paste the ASCII version into
+an email and then you can decrypt it:
+
+\begin{lstlisting}[language=bash]
+$ gpg --decrypt secret.txt.gpg
+
+You need a passphrase to unlock the secret key for
+user: "Ross Fenning <gpg@rossfenning.co.uk>"
+4096-bit RSA key, ID CEEFF134, created 2014-12-14 (main key ID B8D85C76)
+
+gpg: encrypted with 4096-bit RSA key, ID CEEFF134, created 2014-12-14
+      "Ross Fenning <gpg@rossfenning.co.uk>"
+Hello
+$
+\end{lstlisting}
+
+\noident and the ``Hello'' message is displayed. Note that you have to
+provide your passphrase, which means your private key is involved in
+decrypting the message. A malicious party would have to obtain both
+your private key and the passphrase to use it in order to read this
+encrypted message.
 

### 2014-12-15 logbook.tex
Written first lab report.

@@
 
-\title{The Folly of Passwords}
+\title{Practical Security Log Book}
 \author{Ross Fenning}
@@
 
-\chapter{Wireshark Analysis}
+\chapter{Wireshark Analysis of a Teardrop attack}
+
+In this report, we look at a very old attack at the IP layer
+known as the \emph{Teardrop} attack which was able to crash
+Linux and Windows computers at the time of discovery.
+We also look at the protocol violation used to achieve this attack
+in the \emph{Wireshark} analysis tool.
+
+\section{What is a Teardrop attack?}
+
+The Teardrop attack \parencite{center1997cert} was discovered
+in late 1997 as a potential vulnerability in Windows 3.1x, Windows 95 and Windows NT
+systems and versions of Linux prior to versions 2.0.32 and 2.1.63.
+The vulnerability could allow a remote attacker to crash computers
+running these systems due to a bug in the code performing IP fragmentation
+reassembly.
+
+IP fragmentation \parencite{postel1981rfc,fall2011tcp} is necessary if a router
+receives a packet larger than the maximum transmission unit (MTU) of the
+next hop. When this occurs, a router has the option to break the datagram
+into smaller fragments. RFC 815 \parencite{clark1982ip} describes
+some algorithms for how downstream hops can then reassemble the fragmented
+IP packets.
+
+The Teardrop attack involves sending fragmented IP packets such that the offset
+of later packets do not correspond with the length of earlier packets. Normally,
+if the first fragment is 16 bytes in size, then the second fragment should
+indicate an offset of 16 bytes. Setting the second packet to a lower
+offset such that the packets seem to overlap was known to cause bugs in
+these old operating systems and thus they became vulnerable to the Teardrop
+attack.
+
+\section{Example}
+
+Appendix~\ref{appendix:packets} shows raw exports from Wireshark of two frames
+found in a capture of a Teardrop attack. We can see that Wireshark has identified
+frames 8 and 9 as being two IP fragments which together make up a single UDP
+packet.
+
+The IPv4 header advertises a total length for the first fragment (frame 8)
+of 56 bytes. The frame comprises 20 bytes for the IPv4 header and 36 bytes of
+data. The bytes \texttt{00 38} at offset \texttt{0010} show this header length.
+The third bit is set on the bytes \texttt{20 00}, which indicates more
+fragments are to come, but this fragment starts at offset 0 (i.e. it's the first
+fragment). This is a fairly standard initial fragment and there is nothing
+malformed in this frame.
+
+The attack comes in the next fragment (frame 9) where bytes \texttt{00 03}
+claim that this fragment has an offset of 3 eight-byte blocks, or 24 bytes.
+This is not in accord with the fact that the previous fragment was 36 bytes
+long and looks as if this fragment overlaps the previous one.
+
+Wireshark
+has reassembled this to an IPv4 packet with 28 bytes of data as it has
+resolved the malformed fragmentation by losing the last 12 bytes of the
+first fragment (i.e. scrolling back to the advertised offset of 24 and
+then adding on the 4 bytes contained in the second fragment). It has
+also identified that the resulting 28 byte payload length does not
+agree with the 36 byte header claimed in the UDP header (bytes \texttt{00 24}
+at offset \texttt{0026}) and flagged that as an error.
+
+The older versions of Windows and Linux were vulnerable to bugs in
+their fragmentation reassembly that would simply cause the operating
+systems to crash when this overlap occurred.
 
@@
 
-\noident and the ``Hello'' message is displayed. Note that you have to
+\noindent and the ``Hello'' message is displayed. Note that you have to
 provide your passphrase, which means your private key is involved in
@@
 
+\appendix
+
+\chapter{Teardrop Attack Packets}
+\label{appendix:packets}
+
+\lstinputlisting[basicstyle=\scriptsize]{frame8.txt}
+\pagebreak
+\lstinputlisting[basicstyle=\scriptsize]{frame9.txt}
+
 \end{document}

### 2014-12-15 logbook.tex
Finished writing lab 3.

@@
 into smaller fragments. RFC 815 \parencite{clark1982ip} describes
-some algorithms for how downstream hops can then reassemble the fragmented
+an algorithm for how downstream hops can then reassemble the fragmented
 IP packets.
@@
 system although many of the steps apply to other Linux and Unix systems.
-
 The steps outlined below will demonstrate how to:
@@
 \item use the public key to encrypt a plain text message; and
-\item use the private key to descrypt the message.
+\item use the private key to decrypt the message.
 \end{enumerate}
@@
 The gpg tool is normally pre-installed in a lot of Linux/Unix systems as
-tools like \texttt{apt} depends on it. On
+tools like \texttt{apt} depend on it. On
 a Debian or Ubuntu GNU/Linux system that does not have it installed, it
@@
 Note that on the first use of the command, it creates some configuration files
-and keyrings. If there are any keys already stored in the user's home
+and keyrings. If there were any keys already stored in the user's home
 directory, then details of them would be printed out as well.
@@
 you may be asked to generate some entropy.
-Machines had a hard time using a good source of randomness to make sure
+Machines have a hard time using a good source of randomness to make sure
 the key is not predictable and guessable, so you may be required to do
@@
 
-they are then able to encrypt this secret message so that only you may read it:
+\noindent they are then able to encrypt this secret message so that only you may read it:
 
@@
 the encrypted message more suitable for pasting into emails (in a file
-called secret.txt.asc).
+called \texttt{secret.txt.asc}).
 
@@
 
-\chapter{Ethical Hacking Lab Setup}
+\chapter{Ethical Cracking Lab Setup}
+
+In this report, we look at some tools that can be used to put together
+a small lab setup for practising computer cracking and penetration
+testing. This setup could also be used to observe and trace behaviours
+of malware, e.g. viruses, but perhaps with a more stringent approach
+to network isolation (see section~\ref{sec:netsec}).
+
+The setup here is a network of virtual machines running under Virtualbox
+running on a host computer that can be isolated if necessary. Once the
+setup has been designed, we will look at using Vagrant to make a repeatable
+way of bootstrapping the lab on demand. This will help if we ever feel the
+need to tear down the entire setup and start afresh or to share the
+configuration with other researchers.
+
+\section{Kali}
+
+Kali is a GNU/Linux distribution based on Debian that comes with hundreds
+of tools preinstalled that can be used in cracking and penetration testing. It
+is available to download as an ISO CD image file which can easily be used
+to install in a virtual machine using Virtualbox.
+
+The Vagrantfile configuration in appendix~\ref{appendix:vagrant} shows a
+simple provision of a Virtualbox instance of Kali from a pre-built
+Vagrant box for convenience. For more control or more assurance about the
+security of the virtual machine, it should be possible to fork and customise
+the Packer build that created this image.\footnote{https://github.com/ctarwater/packer-kali}
+
+\section{Metasploitable}
+
+Metasploitable is a pre-built virtual machine from Rapid 7 to aid with learning
+its penetration testing tool, Metasploit. We can equally make use of this
+for learning Kali. As with Kali, we can add a configuration to our Vagrantfile
+(appendix~\ref{appendix:vagrant}) for a pre-built box for convenience. Other
+Packer-based builds are available on Github\footnote{https://github.com/waratek/vagrant-boxes}
+if it is again preferred to have more control over the build.
+
+\section{Vagrant}
+
+As discussed in the previous two sections, vagrant provides us two conveniences
+in building our lab setup:
+
+\begin{enumerate}
+\item Repeatable, refreshable and shareable builds of our whole lap setup. A whole classroom
+  of lab setups for teaching could be running within an hour.
+\item The ability to use pre-built box images, at least in the first instance, to get set up
+  quickly, but still migrate to in-house builds of those images when the time is right.
+\end{enumerate}
+
+We can configure vagrant with the \emph{Vagrantfile} shown in appendix~\ref{appendix:vagrant},
+which is a reasonably trivial configuration to bootstrap both Kali and Metasploitable on
+the same private network. Note that the GUI is enabled so we can log in to X or a TTY
+on either machine, which is required because the NAT adapter is overridden such that the
+machines cannot communicate with the outside world. A side-effect of this is that
+vagrant is unable to configure the network interfaces on the guest machines automatically
+for us and we must edit \texttt{/etc/network/interfaces} ourselves, e.g.:
+
+\begin{lstlisting}
+iface eth0 inet static
+    address 192.168.51.5
+    netmask 255.255.255.0
+\end{lstlisting}
+
+The reasoning around this is discussed in section~\ref{sec:netsec}.
+
+\section{Network Security}
+\label{sec:netsec}
+
+By overriding the NAT interface vagrant wants to use, we can enforce that
+our lab network is only using a host-based adapter on a private network such
+that the guests do not even have a gateway to the outside world. We can see
+on the routing table that the guest only knows how to talk to the private
+network:
+
+\begin{lstlisting}[language=bash]
+vagrant@metasploitable:~$ route
+Kernel IP routing table
+Destination     Gateway         Genmask         Flags Metric Ref    Use Iface
+192.168.51.0    *               255.255.255.0   U     0      0        0 eth0
+vagrant@metasploitable:~$
+\end{lstlisting}
+
+For further security, it may be desirable to add iptables rules on the host
+machine to drop any traffic that isn't going between the two defined
+static IPs. This is perhaps an unnecessary step if we are only using the
+penetration and exploit testing tools by hand and are in control of what
+we are doing.
+
+This lab setup can however be extended to include Windows
+machines in the Vagrantfile such that we are able to examine malware and
+viruses. If we are running malicious code (as opposed to simply running
+exploitable services), then we likely want to make sure the host machine
+is simply disconnected from any network and reinstall the whole host
+before it is allowed back on any public network.
 
@@
 
+\chapter{Ethical Cracking Lab Vagrant Configuration}
+\label{appendix:vagrant}
+
+\lstinputlisting[basicstyle=\scriptsize,language=ruby]{networks/pentest/Vagrantfile}
+
 \end{document}

### 2014-12-15 paper.tex
Really trying to cut the word count down.

@@
 \usepackage[style=authoryear,backend=biber]{biblatex}
+\usepackage{rotating}
 
-\title{The Folly of Passwords}
+\title{The Folly of Password Policies: The case for organisations implementing technical alternatives over trying to manage users}
 \author{Ross Fenning}
@@
 that must implement any security processes and policies. In this paper, we
-note that these benefits are primarily for the organisation more than the
+argue that these benefits are primarily for the organisation more than the
 users, so are the usability and effort costs placed on users justifiable
@@
 
-In chapter~\ref{chapter:alternatives}, these alternatives will be
+In chapter~\ref{chapter:alternatives}, some of these alternatives will be
 compared and assessed for their ease of adoption within a business.
@@
 \label{chapter:costs}
-
-The username-password pair is arguably a \emph{de facto} standard
-authentication mechanism employed by IT and security departments
-within many businesses. The default install of any operating
-system (Linux, Mac OS X, Windows, etc.) treats all user accounts
-as consisting minimally of these pair of credentials and a
-username-password login prompt is frequently the first
-screen to greet users when a newly-installed OS is booted
-for the first time.
 
@@
 
-%Linux and Mac systems make use of the Pluggable Authentication
-%Mechanism (PAM), which weakens the assumption that authentication
-%must be done with a password (password authenication is just one
-%module available for PAM). Corporate systems typically make
-%use of centralised user directories (e.g. Active Directory, OpenLDAP),
-%which allows users to use common credentials across multiple
-%IT systems (reducing the need for users to memorise multiple
-%sets of credentials).
-
 \section{Passwords and Usability}
 
-\cite{adams1999users} famously asserted 15 years ago that security
+\textcite{adams1999users} famously asserted 15 years ago that security
 departments take an inappropriate approach to security risk
@@
 
-\cite{herley2009so} argues that users rationally reject costs
+\textcite{beautement2009compliance} later introduced the idea of a
+``compliance budget'' which models how users perceive these costs
+Their work suggests a cap on the effectiveness of any
+security measures within an organisation and reinforce the ideas
+that users will purposefully work around requirements imposed via
+management.
+
+\textcite{herley2009so} argued that users rationally reject costs
 placed upon them from security policies. He invokes Kuhn's
@@
 user base such as employees or indeed the general public, but
-a strong need for true change can be taken from the analogy
+a strong need for change can be taken from the analogy
 at least.
@@
 
-This effect is classed
-by Herley as a negative externality on computer users in
-that they are taking on additional effort to cover costs
-that are not their own. These economic arguments can be
-used to make a case for organisations taking more
-responsibility and investing more to prevent the costs
-from a security compromise. This can be combined with
-the arguments for a more user-centric design
-\parencite{adams1999users} to build a case for a new
-authentication paradigm that is low friction to users and
-where economic incentives lead to an equilibrium where
-organisations spend exactly as much on security as
-is rationally justified by their real risks and costs.
-
-\cite{beautement2009compliance} introduced the idea of a
-``compliance budget'' which models how users perceive these costs
-and benefits. Their work suggests a cap on the effectiveness of any
-security measures within an organisation, but also introduces management
-approaches to raise the threshold above which a user will no longer
-comply with security policies. This arguably increases the cost-benefit
-ratio by making it more worthwhile for users to practice secure
-behaviour (even if by increasing sanctions such that avoidance thereof becomes
-more beneficial than working around costs from complying with security
-requirements). They still admit there is a finite limit to improvements to
-this compliance budget through management, training and cultural changes.
-This supports the idea that many of the costs simply must be moved to the
-organisation itself.
-
-So, what is the nature of the burden experienced by users?
 In the next few sections, we look at various aspects
@@
 a password is to suffer from an attacker learning the password via brute-force
-guessing. A brute-force attack can be said to be \emph{online} (where the
-attacker is sending multiple login attempts directly to a system protecting
-by authentication) or \emph{offline} (where an attacker has a local copy of
-an encrypted file or hashed password, to which she can perform an unlimited
-number of attempts at cracking the password without arousing suspicion).
-
-The advice for creating strong passwords normally takes the form of tips
+guessing. The advice for creating strong passwords normally takes the form of tips
 such as:
@@
 
-This presents a problem if users are poor at working with the
-very kind of passwords needed to avoid brute-force cracking. It
-would suggest that a user-centric design would lead to something
-more in line with abilities humans \emph{do} have.
-
-\textcite{florencio2007large} noted that strong passwords are just
+\textcite{florencio2007large} noted however that strong passwords are just
 as easily stolen through phishing and key logging as weak
-passwords. Password strength is thus limited in only protecting
+passwords. Password strength is thus limited in only protecting against
 brute force attacks and not problems where users accidentally
@@
 is that it likely does not want to suffer
-a breach in an account that had a weak password. Some
+a breach in any account that has a weak password. Some
 online brute force attacks have been known to take place
@@
 is fundementally challenged in the influential findings
-of \textcite{adams1999users}. Their work combined with the
-economics-based arguments and models introduced by %WHO?
-builds a large weight of evidence that well-informed
-users will still not follow much security advice
-(sometimes rationally so) to reduce their own costs in
-time and effort or to remove a significant barrier
-to performing their work tasks.
+of \textcite{adams1999users}. This combined with the
+compliance budget introduced by \textcite{beautement2009compliance}
+suggests management and training only have limited effect.
 
@@
 
-As with password strength, this is only mitigating the
-brute-force guessing attack vector as phishing, keylogging, etc.
-can still give an attacker a password for a short while -- until
-the password is next changed. As with password strength, this
-is also most effective against offline brute-force attacks (e.g.
-the attacker has acquired a list of hashed passwords) an
-account locking policy also prevents an online brute-force
-attack without the need for passwords to expire.
-
 It is a fairly common practice in many organisations to
@@
 
-\textcite{adams1999users} note that password expiry increases
+As with password strength, this is only truly mitigating the
+brute-force guessing attack vector. A password obtained via
+phishing, for example, allows an attacker enough time to do
+real damage before the next password expiration.
+
+\textcite{adams1999users} note also that password expiry increases
 the burden on users having to memorise multiple passwords and that
@@
 by having password strength requirements that are stronger than
-services less concerned with security, with which employeess may also have
+services less concerned with security, with which employees may also have
 accounts. This is susceptible to the issues in section~\ref{sec:strength}
@@
     time and effort.
-  \item Password expiration policies are similar in only protecting against
+  \item Password expiration policies are similar in only really protecting against
     brute-force guessing and their effectiveness is called into question when
@@
 
-In summary, the use of passwords is problematic for organisations and policies
+In summary, the use of passwords is problematic for organisations. Policies
 to reduce those problems place the costs on the wrong people: the users. We can
@@
 In chapter~\ref{chapter:alternatives}, we look at some alternatives to passwords
-and evaluate them.
+and evaluate their suitability for use in the enterprise.
 
@@
 adopting the alternative alongside password authentication in principle. We
-gave some consideration for the ease of doing so in our evaluation. We also
-would advise always doing so as part of a design that considers the users,
-models their use cases and ensures that the approach is indeed increasing
-security more than it is just moving the costs onto the users.
+gave some consideration for the ease of doing so in our evaluation.
 
-\textcite{van2011tiqr} introduced a new alternative authentication mechanism
-called \emph{tiqr} that uses QR codes and a bespoke smartphone app. In
-doing so, he also listed and evaluated a number of existing alternatives. We
-can thus make a working list of existing non-password authentication
-mechanisms:
+We put together a working list of existing non-password authentication
+mechanisms from a combination of the academic literature and the open source
+developer community:
 
@@
   \item TIQR (a web authenication approach that involves QR codes and a
-    bespoke smartphone app)
+    bespoke smartphone app -- \textcite{van2011tiqr})
   \item PAM USB (a module for PAM that turns any USB mass storage device
     into a physical key to unlock or lock a computer)
+  \item YubiKey is a bespoke hardware device capable of One-Time Passwords (OTP),
+    storing static passwords and also challenge-response.
 \end{itemize}
 
-PAM USB\parencite{steele2005paranoid} is an additional mechanism added to
+PAM USB \parencite{steele2005paranoid} is an additional mechanism added to
 this list. It allows users of Unix systems to log in by plugging in a USB
 mass storage device.
-
-In this chapter, we will explore the feasibiity of adopting some of
-these authentication methods within an enterprise. The primary consideration
-is around detailing the setup and maintenance steps and whether any
-method can be justifiably ruled out in the early phases of a feasibility
-study. In section~\ref{sec:evaluation}, we will look at how the evaluation
-was carried out and the remaining sections detail that evaluation against each
-considered method.
 
@@
 evaluated for use in an organisational setting. The focus of this evaluation
-was to detail the steps involved in creating a toy network of machines
+was to run through the steps involved in creating a toy network of machines
 that allowed common authentication across a representative sample of
@@
     being evaluated;
-  \item \textbf{Company Email}: users can then access their corporate email without needing additionally
+  \item \textbf{Company Email}: users can then access their corporate email without additionally
     supplying a password to do so;
@@
     servers (this assumes a company where this is a common work task, e.g.
-    software development, which is including to exercise the authentication
+    software development, which is included to exercise the authentication
     further rather than to act as a typical business use case).
@@
 companies to have an authoritative directory server so that users
-can at least use the same password across all company services. Some businesses
+can at least use the same password across all company services.
+
+Some businesses
 attempt true SSO and ensure users only type their password during initial
 login and they are automatically authorised against all revelant systems
-thereafter. This is recommended practice to reduce some of the effort
-remembering multiple passwords or even typing the same password several times
-a day. Ideally, a suitable password alternative would provide similar
+thereafter.
+
+Ideally, a suitable password alternative would provide similar
 convenience in minimising effort from users when allowing access to a multitude
@@
     ruled out for consideration as it requires mobile operators to install
-    keys on their customers SIM cards, which has been problematic in the
-    past.
-  \item PAM USB.
-  \item YubiKey. This is a hardware USB device from Yubico that can act
-    as an OTP generator, a challenge-response token and even a store
-    for a static password.
+    keys on their customers SIM cards.
+  \item \texttt{pam\_usb}. This module receives little attention in the literature
+    but is a working tool used by many Linux users.
+  \item YubiKey. This device seems promising with its ability to do different
+    authentication mechanisms.
 \end{itemize}
@@
 Another option is to try to centralise the OTP authentication.
-Solutions such as LinOTP suggest the abiity to have a centralised
+Solutions such as LinOTP suggest the ability to have a centralised
 OTP-based authentication system with a custom PAM module. Combined
@@
 including for access to email and web-based applications.
-
-A centralised solution is more flexible for corporate systems
-as PAM does not provide an intrinsic method for fowarding local
-authentication on to other services. Microsoft Windows achieved
-this in the past with NTLM -- which is possible to configure
-also on Unix systems with Winbind -- but that is largely
-regarded as an insecure mechanism today.
 
@@
 
-Some desktop environments (e.g. Gnome, Mac OS X) provide
-a local keychain that can store SSH keys, SSL certificates, etc.
-which are unlocked whilst the user is logged in, but inaccessible
-otherwise. This already provides a convenience desktop
-users, but it relies on the user authenticating with a password
-that is then the same password used to update the keychain
-(Mac OS X updates the keychain password when the user's main
-account password is changed). There maybe be some scope for
-further research into whether dekstop keychains can be made
-compatible with OTP authentication.
-
 There is certainly scope for further work to evaluate extensive
-use of LinOTP and Kerberos within an enterprise setting and the
-benefits they bring over relying on a local desktop keychain.
+use of LinOTP and Kerberos within an enterprise setting as least as
+a case study.
 
@@
 
-The PAM USB module for the PAM system is similar in many
+The \texttt{pam\_usb} module for the PAM system is similar in many
 ways to the OATH OTP module in terms of setup. The module
@@
 drivers removing their keys when they leave their car
-unattended, so this is more in line with existing user
+unattended, which fits with existing user
 habits than asking users not to disclose or to strengthen
@@
 is not easily forwarded to other services on the network.
-
-%An advantage this module has over the OATH OTP module is that
-%local applications can make authentication requests
-%multiple times without the user needing to be involved. As
-%long as the USB device remains connected, applications
-%can authenticate multiple times, but OTP authentication
-%will require a new password to be entered on each authentication
-%request.
 
@@
 
-The use of a PIN does not eliminate the ``something you
-know'' factor, but a short PIN is far less of a burden
-than the requirement for users to have strong, random
-passwords. This could be a reasonable compromise for an
-enterprise given some further usability research.
-
 The difficulty of eliminating a password or PIN from the login workflow
@@
 
-The \emph{Scute} project provides a PKCS~\#11 module that allows an OpenPGP
-smartcard to be used with the Network Security Services (NSS) within
+The \emph{Scute}\footnote{http://www.scute.org/}
+project provides a PKCS~\#11 module that allows an OpenPGP
+smartcard to be used with the Network Security Services (NSS) and
 the Firefox web browser (we believe Chrome also supports this, but Firefox
 was used in our test network). Combined with Apache's \emph{mod\_nss} (which
-derives from mod\_ssl), we were able to implement an authentication layer
+derives from \texttt{mod\_ssl}), we were able to implement an authentication layer
 over a web-based application that supported an OpenPGP smartcard.
@@
 
-Table~\ref{table:overview} summarises the approaches tried on our test network
+Table~4.1 summarises the approaches tried on our test network
 and discussed in chapter~\ref{chapter:alternatives}. In some cases, there
@@
 
-\begin{table}[h]
-  \begin{tabular}{|l|l|l|l|l|} \hline
+\begin{sidewaystable}[p]
+\centering
+  \begin{tabular}{|l|p{4cm}|p{4cm}|p{4cm}|p{4cm}|} \hline
                             & \textbf{OTP Apps} & \textbf{pam\_usb} & \textbf{OpenPGP smartcard} & \textbf{YubiKey} \\ \hline
@@
   \label{table:overview}
-\end{table}
+  \caption{Summary of technologies involved in achieving enterprise requirements from each authentication mechanism.}
+\end{sidewaystable}
 
@@
 
-Thus we can recommend coporations are able to evaluate the hardware costs
+Thus we can recommend corporations are able to evaluate the hardware costs
 and consider YubiKeys as a ``drop in'' replacement for users memorising
@@
 \begin{itemize}
-\item What is the effectiveness and what potential issues are there around
+\item What is the effectiveness of and what potential issues are there around
   a real deployment of LinOTP within a business? An empirical study in a

### 2014-12-15 paper.tex
More trimming

@@
 ``Paradigm Shift''\parencite{kuhn1962structure} to suggest
-that the increasing burden placed on users with every attack
-that takes place and the additional complexity that comes
-therewith may hint at an imminent, yet dramatic change in
-how users are asked to authenticate. It is yet to be seen
-if Kuhn's observations of sudden shifts of thinking
-in the scientific community apply to a more general
-user base such as employees or indeed the general public, but
-a strong need for change can be taken from the analogy
-at least.
+an imminent, yet dramatic change in
+how users are asked to authenticate as the complexity of
+rules placed on users increases with each attack.
 
@@
 from exposure to those costs themselves.
-
-In the next few sections, we look at various aspects
-of password security, how users are normally advised
-and critically examine the effectiveness of that advice.
 

### 2014-12-15 paper.tex
Cut a bit more...

@@
 Passwords are not the only approach to verifying a user's identity.
-The methods by which a user can be authenticated are known to fall
-into three categories -- or \emph{factors} -- of authentication
-\parencite{council2005authentication}:
-
-\begin{itemize}
-  \item something the user \emph{knows}, such as a password, a PIN or
-    personal information (e.g. mother's maiden name);
-  \item something the user \emph{has}, such as a passport, a mobile device
-    a paper ticket; and
-  \item something the user \emph{is}, such as their DNA sequence,
-    their retinal pattern, their fingerprint.
-\end{itemize}
-
-It should be noted that some methods can be argued as spanning
-over two factors. For example, a credit card is something
-a bank customer possesses, but the card number is something
-an attacker could learn without the owner losing ownership
-of the card. In the context of many online purchase, credit
-card \emph{details} arguably become something a user \emph{knows}.
-
 There
@@
 such as Google, Amazon, Microsoft, Paypal and Twitter having enabled
-two-factor authentication for user accounts.
+two-factor authentication for user accounts. Examples include One-Time
+Passwords or challenge-response codes.
 
@@
 
-Brute-force password guessing can also be mitigated by
-users changing passwords frequently enough that attackers
-have a short window within which to guess a particular
-password.
-
 It is a fairly common practice in many organisations to
@@
 Some of the identified alternatives to password authentication were
-evaluated for use in an organisational setting. The focus of this evaluation
-was to run through the steps involved in creating a toy network of machines
-that allowed common authentication across a representative sample of
-classes of services that may be used by businesses.
-
+evaluated for use in an organisational setting.
 The requirements for our toy/proof-of-concept systems were:
@@
     servers (this assumes a company where this is a common work task, e.g.
-    software development, which is included to exercise the authentication
-    further rather than to act as a typical business use case).
+    software development)
 \end{itemize}

### 2014-12-15 paper.tex
Cut even more words out.

@@
 
-These authentication requirements are largely driven by
-several desires:
-
-\begin{itemize}
-  \item to prevent use of systems by non-employees;
-  \item to maintain an audit trail of employees' usage;
-  \item to protect company secrets and intellectual property;
-  \item to safeguard the personal information of customers and employees; and
-  \item to prevent economic and reputational damage from would-be attackers.
-\end{itemize}
-
 Additional security in any system normally comes with additional costs
@@
 argue that these benefits are primarily for the organisation more than the
-users, so are the usability and effort costs placed on users justifiable
-if they are to gain very little?
+users. This calls into question the justifiability of placing
+effort and usability costs on users when they gain little out of it
+personally.
 
@@
 \textcite{herley2009so} argued that users rationally reject costs
-placed upon them from security policies. He invokes Kuhn's
-``Paradigm Shift''\parencite{kuhn1962structure} to suggest
+placed upon them from security policies. He invoked Kuhn's
+``Paradigm Shift'' \parencite{kuhn1962structure} to suggest
 an imminent, yet dramatic change in
@@
 \label{sec:strength}
-
-Password-protected accounts are vunerable to at least three
-classes of attack \parencite{florencio2014password}:
-
-\begin{itemize}
-  \item Malware infection or other direct compromise of the client machine (e.g. keyloggers)
-  \item Attacker learns the passwords via phishing, brute-force guessing or other means
-  \item Access to a particular system is gained without the actual password (e.g session hijacking, cross-site request forgery)
-\end{itemize}
 

### 2014-12-15 paper.tex
Severe cuts being made now.

@@
 
-One classical solution to weak passwords
-has been to enforce password strength \emph{requirements}.
+Many companies enforce password strength \emph{requirements}.
 This is a technical solution whereby the systems simply
@@
 to a target strength or entropy.
-This is a technical solution that shifts the burden forcibly onto
+This shifts the burden forcibly onto
 users and we can argue it is susceptible to the deliberate workarounds
@@
 
-There is certainly scope for further work to evaluate extensive
-use of LinOTP and Kerberos within an enterprise setting as least as
-a case study.
-
 \section{PAM USB}
@@
 
-Futher research may find a solution to storing keys
-and certificates on a USB device without passwords or highlight
-whether the risks of doing so are too high.
-
 \section{PKI Tokens and Smartcards}
@@
 
-The difficulty of eliminating a password or PIN from the login workflow
-suggests that the PKI token or smartcard approach is more suitable for
-situations where security requirements are higher and two-factor authentication
-is preferable. The same technology is used for debit/credit card transactions
-and for withdrawing cash from a bank account and thus we would expect it
-to be used in other organisations that have similar security requirements.
-
 The \emph{Scute}\footnote{http://www.scute.org/}
@@
 Given the wide variety of
-options for integrating with services such as email, we did not go as far
-as to evaluate them all and a thorough examination of a smartcard-based,
+options for smartcards, readers and supporting software, we were unable
+to evaluate them all and a thorough examination of a smartcard-based,
 enterprise computer network is a subject for further research. The most
 subtantial work in this area (that we know of) is from \textcite{grossesecuring}.
-This research compares different smartcard technologies and software projects
+This research compared different smartcard technologies and software projects
 and evaluates them for functionality that overlaps with the enterprise
@@
 one-time password for the user, which allows not only for more secure tokens,
-but also increases the overhead of the login process. The YubiKey achieves
-this by acting as a USB keyboard during this operation.
+but also increases the overhead of the login process.
 
@@
 
-A drawback is that the YubiKey lacks additional storage for other certificates
-or keys (as is the case with a USB storage device or an OpenPGP smartcard). This
-makes it difficult to forward the authentication to remote systems. Yubico
-produce the more expensive YubiKey Neo which supports OpenPGP and acts as
-a smartcard. We did not evaluate this model in our test network.
-
 Perhaps the most promising feature of the YubiKey for business use is the
-ability to store a static password as well as the functionality above. A
-managed distrubution of YubiKey hardward to employees within an organisation
-could involve setting up each key with a strong, static password. The users
-would not need to memorise the password, but it could be used as part of
-a shared-credentials setup (e.g. LDAP directory service already established
-within the organisation) without any significant changes to the technology
-used.
-
-In chapter~\ref{chapter:conclusion}, we explore the implication of the
-static password feature within a company and draw some conclusions about the
-different password alternatives and their use in the enterprise.
+ability to store a static password as well as the functionality above. This
+leads to the possibility of introducting the device in a way that is
+compatible with exist applications that require a username and password.
 

### 2014-12-15 paper.tex
Tidied up log book

@@
 
-The challenge here is that each user needs a unique secret which has to
-be managed on a trusted server. The secret for every employee then needs
-to be available in a configuration file on every desktop computer. This
-can be achieved with configuration management systems such as Puppet,
-Chef, Ansible, Salt, etc.
-
-Another option is to try to centralise the OTP authentication.
 Solutions such as LinOTP suggest the ability to have a centralised

### 2016-02-20 logbook.tex
Tidied up log book

@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ sudo apt-get install gnupg
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --list-keys
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --gen-key
@@
 some work via keystrokes, mouse movements or run some processes that
-do disk activity. Once this is complete, you will have your public and
+do disk activity to generate some random events from which entropy
+can be derived. Once this is complete, you will have your public and
 private key pair.
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --list-secret-keys
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --output mypublickey.gpg --export B8D85C76
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --armor --export B8D85C76
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --import mypublickey.pgp
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --fingerprint B8D85C76
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --sign-key B8D85C76
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --output B8D85C76-signedby-12345678.gpg  --export B8D85C76
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --import B8D85C76-signedby-12345678.gpg
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ echo "Hello" >secret.txt
@@
 
-\begin{lstlisting}[language=bash]
-$ echo "Hello" >secret.txt
-$ cat secret.txt 
-Hello
-\end{lstlisting}
-
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --recipient B8D85C76 --encrypt secret.txt
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 $ gpg --decrypt secret.txt.gpg
@@
 
-The setup here is a network of virtual machines running under Virtualbox
+The setup here is a network of virtual machines running under Virtualbox\footnote{https://www.virtualbox.org/}
 running on a host computer that can be isolated if necessary. Once the
-setup has been designed, we will look at using Vagrant to make a repeatable
+setup has been designed, we will look at using Vagrant\footnote{https://www.vagrantup.com/}
+to make a repeatable
 way of bootstrapping the lab on demand. This will help if we ever feel the
@@
 
-Kali is a GNU/Linux distribution based on Debian that comes with hundreds
+Kali\footnote{https://www.kali.org/}
+is a GNU/Linux distribution based on Debian that comes with hundreds
 of tools preinstalled that can be used in cracking and penetration testing. It
@@
 
-Metasploitable is a pre-built virtual machine from Rapid 7 to aid with learning
+Metasploitable\footnote{http://sourceforge.net/projects/metasploitable/}
+is a pre-built virtual machine from Rapid 7 to aid with learning
 its penetration testing tool, Metasploit. We can equally make use of this
@@
 
-\begin{lstlisting}[language=bash]
+\begin{lstlisting}
 vagrant@metasploitable:~$ route
@@
 
+\printbibliography
+
 \appendix
